---
title: "Quiz 1"
author: "Anna Miletova, 89231151"
date: "2025-03-23"
output: 
  pdf_document:
    latex_engine: xelatex
---

```{r}
my.data <- read.table("data.txt", header = TRUE, sep = "\t")

x <- 1
z <- 5

my.data <- my.data[unique(c(seq(x,nrow(my.data),by=10),seq(z,nrow(my.data),by=10))),]
```

## 1. Histogram for Books..how.many.per.year

```{r}
books <- my.data$Books..how.many.per.year

hist(books, main="Histogram of Books", xlab="Books per Year")
```

The histogram is positively asymmetric.

Since the graph is asymmetric to the right we expect median to be more than average and median to be a more appropriate measure.

### Central tendency measures

```{r}
modus <- function(v) {
  uniq_vals <- unique(v)
  uniq_vals[which.max(tabulate(match(v, uniq_vals)))]
}

cat(
  "Average: ", mean(books), 
  "\nMedium: ", median(books), 
  "\nModus: ", modus(books)
)
```

### Measures of variability

```{r}
# cat("IQR: ", quantile(books, 0.75) - quantile(books, 0.25))
cat(
  "Range: ", max(books) - min(books), 
  "\nIQR: ", IQR(books), 
  "\nStandard deviation: ", sd(books)
)
```

## 2. Boxplot for Friends.on.Facebook

```{r}
friends <- my.data$Friends.on.Facebook

boxplot(friends, main="Boxplot of Facebook Friends", ylab="Friends")
```

```{r}
summary(friends) # Min, Max, Mean, Median, Quartiles
IQR(friends, na.rm=TRUE)
```

The outliers, points outside of the "whiskers", are not observed in the boxplot.

## 3. Circumference of a circle with a radius r

```{r}
circumference <- function(r) {
  return(2 * pi * r)
}
cat("Circumference of the circle, where r = x =", x ,"is", circumference(x))
```

## 4. Roulette probability

```{r}
times <- as.numeric(paste0(x, 5))
win_probability <- 1/37

P_X_equals_z <- dbinom(z, size = times, prob = win_probability)
cat("P(X =", z, ") =", P_X_equals_z)
```

### PDF

```{r}
k_values <- 0:times
pmf_values <- dbinom(k_values, size = times, prob = win_probability)

plot(
  x    = k_values, 
  y    = pmf_values, 
  type = "s",
  main = "PMF of X ~ Binomial(15, 1/37)",
  xlab = "Number of wins (k)",
  ylab = "P(X = k)"
)
```

### CDF

```{r}
cdf_values <- pbinom(k_values, size = times, prob = win_probability)

plot(
  x    = k_values, 
  y    = cdf_values, 
  type = "s",
  main = "CDF of X ~ Binomial(15, 1/37)",
  xlab = "Number of wins (k)",
  ylab = "P(X <= k)"
)
```

### Probability of winning less than 5 times

```{r}
P_less_than_5 <- pbinom(4, size = times, prob = win_probability)
cat("P(X < 5) =", P_less_than_5)
```
