---
title: "Quiz 3"
author: "Anna Miletova, 89231151"
date: "2025-05-22"
output: 
  pdf_document:
    latex_engine: xelatex
---

```{r}
library(pwr)

my.data <- read.table("data.txt", header = TRUE, sep = "\t")

x <- 1
z <- 5

my.data <- my.data[unique(c(seq(x,nrow(my.data),by=10),seq(z,nrow(my.data),by=10))),]
```

## Task 1. Check shoe size difference for males and females

```{r}
t.test(Shoe.size ~ Sex, data = my.data, var.equal = TRUE, conf.level = 0.99)
pwr.t.test(d = 1.5 / 2, n = 25, sig.level = 0.01, type = "two.sample")
```
Since p-value is significantly less than 0.01, we reject H0 null hypothesis and therefore shoe size for male and female is significantly different.

The calculated power of the two-sample t-test is approximately 0.49, which is below the commonly recommended threshold of 0.80. This means that there is a less than 50% chance of correctly detecting a true difference in shoe size between males and females if such a difference exists.

## Task 2. Graph of Weight vs Shoe Size

```{r}
plot(x=my.data$Shoe.size, y=my.data$Weight,
     xlab = "Shoe Size", ylab = "Weight (kg)",
     main = "Weight vs Shoe Size")
```

The plot reveals a positive trend — as shoe size increases, weight also tends to increase. However, the relationship is not perfectly linear, and there is considerable spread in the data for each shoe size, suggesting moderate variability.

## Task 3. Weight and Sport Hours Per Week correlation

```{r}
cor.test(my.data$Weight, my.data$Sport..hours.per.week., method = "spearman")
```

Since p-value < 0.05, we reject the H0 null hypothesis. The value ρ = 0.295 indicates a moderate positive relationship. This suggests that, in this sample, students who report spending more time doing sports also tend to weight more.

There is a statistically significant monotonic relationship between weight and sport hours per week in the population.

## Task 4. Pet owners ratio check

To check if the ratio is different than dog 41%, cat 33%, fish 10%, other 16%.

```{r}
dog_count <- sum(my.data$PetDog == "Yes")
cat_count <- sum(my.data$PetCat == "Yes")
fish_count <- sum(my.data$PetFish == "Yes")
other_count <- sum(my.data$PetOther == "Yes")
observed <- c(dog_count, cat_count, fish_count, other_count)
total_count <- sum(observed)

expected.proportions <- c(0.41, 0.33, 0.10, 0.16)
expected <- total_count * expected.proportions

chisq.test(observed, p = expected.proportions)
```


Since the p-value is much greater than 0.05, we fail to reject the null hypothesis.
There is no significant difference between the observed pet ratio and the expected one.
