---
title: "Quiz 2"
author: "Anna Miletova, 89231151"
date: "2025-04-26"
output: 
  pdf_document:
    latex_engine: xelatex
---

```{r setup, include=FALSE}
# knitr::opts_chunk$set(echo = TRUE)
library(dplyr)
```

```{r}
my.data <- read.csv("data.txt", sep="\t")

x <-5
z <-1
my.data <- my.data[unique(c(seq(x,nrow(my.data),by=10),seq(z,nrow(my.data),by=10))),]
```

## Task 1. Check if he population mean of male students is greater than 180 cm.

Null hypothesis: the mean height of male students is 180 cm or less.

Alternative hypothesis: the mean height is greater than 180 cm.

```{r}
male.data <- my.data %>%
  filter(Sex == "male") %>%
  pull(Height)

t.test(male.data, mu = 180, alternative = "greater")
```

Since we take α = 0.05 and p-value is 0.0009434 \< 0.05, we reject the null hypothesis and make a conclution that the mean height of male students is greater than 180 cm.

## Task 2. Plot and distribution.

```{r}
hist(male.data, 
     main = "Histogram of Male Heights", 
     xlab = "Height (cm)",
     breaks = 10)
```

Histogram shows that the data from previous exercise is not perfectly normally distributed.

## Task 3.

```{r}
p <- 0.5
n <- 30
num_samples <- 200

sample_means <- numeric(num_samples)

xbar <- replicate(num_samples, mean(rgeom(n, p)))
hist(xbar,
     breaks = 20,
     main   = expression(paste("Sampling distribution of ", bar(X)[n])),
     xlab   = expression(bar(X)[n]))
```

Distribution is close to normal.

## Task 4.

```{r}
power.t.test(delta = 25, sd = 50,
  sig.level = 0.05, power = 0.8,
  type = "paired", alternative = "two.sided")
```

n = 33.3672

Running a power analysis for a paired-sample t-test (α = 0.05, power = 0.80, aiming to spot a 25 mg/dL drop with 50 mg/dL variability) shows we need about 34 subjects. That’s the point where we’re likely (80 %) to catch a real effect.

```{r}
n <- 50
before <- rnorm(n, mean = 150, sd = 40)
after <- before - rnorm(n, mean = 25, sd = 50)

head(cbind(before, after), 6)

test <- t.test(before, after, paired = TRUE)
print(test)


cohen_d <- with(test, estimate / (sd(before - after)))
cohen_d
```

With our sample of 50 paired readings, the paired t-test gave a p-value \< 0.001 and an average drop of 24 mg/dL, which means that blood-sugar levels went down in a statistically meaningful way; therefore treatment works.
